### 2017-11-15 submission/manuscript.Rmd
update fig4 and fig5 legend
nick's abstract suggestions

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC). Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

@@

**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only. B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only. The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.

**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only. B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only. The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.

>>>>

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced. Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples. For the included studies, most were adequately powered for large effect size differences. This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.

@@

**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only. B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only. The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between. The solid red line represents the median AUC of all the studies for either the full community or CRC-associated genera OTUS only model.

**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only. B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only. The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between. The solid red line represents the median AUC of all the studies for either the full community or CRC-associated genera OTUS only model.

### 2017-11-15 submission/manuscript.tex
update fig4 and fig5 legend
nick's abstract suggestions

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer (CRC). Important
drivers within this context have ranged from individual microbes to the
whole community. Our study expands on a recent meta-analysis
investigating microbial biomarkers for CRC by testing the hypothesis
that the bacterial community is an important driver of both early
(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
we examined both feces (n = 1737) and tissue (492 total samples from 350
individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in a higher classification
based on Area Under the Curve (AUC) for both adenoma and carcinoma using
fecal and tissue samples. For the included studies, most were adequately
powered for large effect size differences. This may be more amenable for
carcinoma than for adenoma microbiota research due to the smaller
community level changes observed in our results.

@@

\textbf{Figure 4: OTU Random Forest Model of Stool Across Studies.} A)
Adenoma random forest model between the full community and
CRC-associated genera OTUs only. B) Carcinoma random forest model
between the full community and CRC-associated genera OTUs only. The
dotted line represents an AUC of 0.5 and the lines represent the range
in which the AUC for the 100 different 80/20 runs fell between.

\textbf{Figure 5: OTU Random Forest Model of Tissue Across Studies.} A)
Adenoma random forest model between the full community and
CRC-associated genera OTUs only. B) Carcinoma random forest model
between the full community and CRC-associated genera OTUs only. The
dotted line represents an AUC of 0.5 and the lines represent the range
in which the AUC for the 100 different 80/20 runs fell between.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for CRC by
testing the hypothesis that the bacterial community is an important
driver of both early (adenoma) and late (carcinoma) stage of disease. To
test this hypothesis we examined both feces (n = 1737) and tissue (492
total samples from 350 individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness (P-value
\textless{} 0.05) after correcting for study effect and variable region
sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
in a slightly increased relative risk while lower Shannon diversity and
evenness in fecal samples resulted in a significant increase in relative
risk for carcinoma (P-value \textless{} 0.05). Previously associated
colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
pattern with a significantly increased relative risk by their presence
for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
\textless{} 0.05). Using the whole community versus only CRC associated
genera to build a prediction model resulted in higher classification
success based on Area Under the Curve (AUC) for both adenoma and
carcinoma using fecal and tissue samples. For the included studies, most
were adequately powered for large effect size differences. This may be
more amenable for carcinoma than for adenoma microbiota research due to
the smaller community level changes observed in our results.

@@

\textbf{Figure 4: OTU Random Forest Model of Stool Across Studies.} A)
Adenoma random forest model between the full community and
CRC-associated genera OTUs only. B) Carcinoma random forest model
between the full community and CRC-associated genera OTUs only. The
dotted line represents an AUC of 0.5 and the lines represent the range
in which the AUC for the 100 different 80/20 runs fell between. The
solid red line represents the median AUC of all the studies for either
the full community or CRC-associated genera OTUS only model.

\textbf{Figure 5: OTU Random Forest Model of Tissue Across Studies.} A)
Adenoma random forest model between the full community and
CRC-associated genera OTUs only. B) Carcinoma random forest model
between the full community and CRC-associated genera OTUs only. The
dotted line represents an AUC of 0.5 and the lines represent the range
in which the AUC for the 100 different 80/20 runs fell between. The
solid red line represents the median AUC of all the studies for either
the full community or CRC-associated genera OTUS only model.
